---
title: "Presidential Campaign Contributions in Alabama 2016"
author: "Jacob Dodd"
date: "11/19/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Measuring Presidential Campgain Contributions from the 2016 cycle from Alabama

I seek to analyze campgain contributions to different presidential campgaingns diring the 2016 presidential election cycle from Alabama. Data was downloaded from http://classic.fec.gov/disclosurep/PDownload.do. 

Importing relevant Libraries
```{r}
rm(list=ls())
setwd("/Users/jacobdodd/Desktop")

library(ggplot2)
library(dplyr)
library(reshape2)
library(data.table)
library(lubridate)
library(knitr)
library(tidyr)

```

### Loading the dataset and first look
```{r}
AL <- fread("ALP2016.csv",sep = ",",data.table = TRUE,stringsAsFactors = FALSE)

#Starting data input on line 2 and discarding line 1 because it has too few or too many items to #be column names or data: 
names = c("cmte_id","cand_id","cand_nm","contbr_nm","contbr_city","contbr_st","contbr_zip","contbr_employer","contbr_occupation","contb_receipt_amt","contb_receipt_dt","receipt_desc","memo_cd","memo_text","form_tp","file_num","tran_id","election_tp")

#fread mistakenly loaded a 19th column when there are only 18, so I have to drop it before reassigning variable names.
AL$V19 <- NULL

#reassigning names
colnames(AL) <- names
rm(names)
str(AL)
```

# Data Review
Some variables that jumped out to me:
- Committee ID
- Candidate ID
- Candidate Name
- Conbritor City
- Contrib. Zip
- Contrib. Occupation
- Receipt Amount
- Recepit Date
- Election Type

It will interesting to do some exploration of giving amounts and counts over time, as well as some spatial potting. Another thing that could be possible is analyze different attributes (city, occupation, etc.) of contributors, and see if there are any common features. 

There are some columns which may not provide much insight:
-receipt description
- memo code
- memo text
- contbr state (all values are from AL)
- contrib. employer (I am doubtful there are enough insightful values for each of the 55000 rows)
- Contr. Name (these should be used to create IDs, then be removed, to de-identify the rows) If anything, I can use these to  create a contributor ID to do some analysis on how many contributions the average Alabaman gave
- form type (not sure what this represents)
- file number(not sure what this represents)


### - Dataset dimensions and auditing
```{r}
print(paste0(nrow(AL)," rows, ",ncol(AL)," columns"))
```
 

### - Auditing columns for NAs
```{r}
for (i in colnames(AL)[1:18]) {
  print(paste0(i,": ",sum(is.na(AL[[i]]))))
}

```
There are few missing values for the columns here, the only substantial are in contributor employer. 


### - Auditing column values
Let's now look at individual column values with the table() function.
#### Candidate information Columns 
```{r}
#Committe ID, Candidate ID, and Candidate Name
for (i in colnames(AL)[1:3]){
  print(table(AL[[i]]))
  print(length(unique(AL[[i]])))
}
```

There were 23 candidates that recieved donation in this election cycle in Alabama. This database does no contain any information on candidate political party or gender. This will be added shortly.

A caveat here is this candidate information only represents candidates that recieved at least one donation, so does not serve as a list of all candidates that ran for president in this cycle. 

From a cursory look, looking at the candidates names via the format here is unwieldy. Let's use the last name for interpretability when doing EDA.

For most of the candidates listed here, I am able to ID which party they ran in. Apparently Lessig and McMullin ran as a Democrat and Independant, respectively. 

#### Creating party name vector containing candidates
```{r}
#adding political party for candidates
republican <- c("Bush, Jeb","Carson, Benjamin S.","Christie, Christopher J.","Cruz, Rafael Edward 'Ted'","Fiorina, Carly","Jindal, Bobby","Huckabee, Mike","Graham, Lindsey O.","Kasich, John R.","Perry, James R. (Rick)","Rubio, Marco","Santorum, Richard J.","Trump, Donald J.","Walker, Scott","Paul, Rand")

democrat <- c("Clinton, Hillary Rodham","Sanders, Bernard","Webb, James Henry Jr.","O'Malley, Martin Joseph","Lessig, Lawrence")

independ <- c("McMullin, Evan")

libert <- c("Johnson, Gary")

green <- c("Stein, Jill")

#checking lengths so each candidate is reprented
sum(length(republican),length(democrat),1,1,1)
```

#### Assigning Party IDs to Candidates
```{r}
#creating party IDs
AL$repub <- ifelse(AL$cand_nm %in% republican, 1, NA)
AL$demo <- ifelse(AL$cand_nm %in% democrat, 2, NA)
AL$green <- ifelse(AL$cand_nm %in% green, 3, NA)
AL$libert <- ifelse(AL$cand_nm %in% libert, 4, NA)
AL$ind <- ifelse(AL$cand_nm %in% independ, 5, NA)

#creating a column to add new values to
AL$party <- NA

#merging columns values back
for (i in colnames(AL)[19:23]) {
  AL$party[!is.na(AL[[i]])] = AL[[i]][!is.na(AL[[i]])]
}

#re-assigning as numeric values
AL$party <- as.numeric(AL$party)

#dropping the unneeded columns
for (i in colnames(AL)[19:23]) {
  AL[[i]] <- NULL
}

```

#### Assigning Candidate Genders
```{r}
#creating vector with female candidate names
female <- c("Fiorina, Carly","Clinton, Hillary Rodham","Stein, Jill")

#assigning gender IDs
AL$cand_gdr <- ifelse(AL$cand_nm %in% female,0,1)
```

#### cleaning candidate names
```{r}
#Let's use the candidate's last name for interpretability in EDA
AL$cand_nm <- vapply(strsplit(AL$cand_nm,","), `[`, 1, FUN.VALUE=character(1))

```

#### Dropping columns with not much descriptive info
```{r}
AL_keep <- c("cmte_id",
             "cand_id",
             "cand_nm",
             "contbr_nm",
             "contbr_city",
             "contbr_zip",
             "contbr_employer",
             "contbr_occupation",
             "contb_receipt_amt",
             "contb_receipt_dt",
             "receipt_desc",
             "memo_text",
             "form_tp",
             "election_tp",
             "party",
             "cand_gdr"
             )

AL <- subset(AL, select = AL_keep)

```

#### Cleaning the Date column
```{r}
AL$contb_receipt_dt <- as.Date.character(AL$contb_receipt_dt,format = "%d-%B-%y")
```

#### Changing the nme column to unique IDs
```{r}
AL <- transform(AL, nameID=match(contbr_nm, unique(contbr_nm)))
AL$contbr_nm <- NULL
```





## Distribution of Donations
#### Donation Summary
```{r}
summary(AL$contb_receipt_amt)
```
A min of -$7300? How is this possible.
```{r}
subset(AL, AL$contb_receipt_amt == min(AL$contb_receipt_amt))
subset(AL, AL$nameID == 5151)

```


Why are there negative donations? Let's explore this further.
```{r}
subset(AL,AL$contb_receipt_amt < 0)
```
There are 935 (~1.7%) negative contributions. It is possible that these are refunds. Let's look at the values with table().
```{r}
negAL <- subset(AL, AL$contb_receipt_amt < 0)
table(negAL$receipt_desc)
```
For the majority of these, there is no description. Others were redesignated to presidential campaigns (Hilary Victory Fund, redisgnation). These will be adjusted to have positive values, and the term type changed to G2016 (for presidential contributions). 
```{r}
#creating a vector for redesignation to general 
general <- c("redesignation to general","* hillary victory fund","redesignation to presidential general")

#lower casing characters
AL$memo_text <- sapply(AL$memo_text, tolower)
AL$receipt_desc <- sapply(AL$receipt_desc, tolower)

#changing election types to general
AL$election_tp[(AL$contb_receipt_amt < 0) & ((AL$memo_text %in% general) | (AL$receipt_desc %in% general))] <- "G2016"

#separating out the the values to make positive
#AL[(AL$contb_receipt_amt < 0) & ((AL$memo_text %in% general) | (AL$receipt_desc %in% general))] <- unlist(lapply(AL$contb_receipt_amt[(AL$contb_receipt_amt < 0) & ((AL$memo_text %in% general) | (AL$receipt_desc %in% general))], abs))

#AL <- unnest(AL, contb_receipt_amt)

```


Others were "reattributed" to spouse, so we must check for duplicate records. 
For the remaining, they will be removed from the dataset, as there is no explanation, most likely being refunds. 

Let's look at one of these contributors donation history.
```{r}
subset(AL,AL$contbr_nm == "RADCLIFF, MARY ROSS")
```
From what we see above, the negative contributions occur after the date of initial contributions, and are in the same amount as the prior contributions. This supports my theory that the unmarked donations are requested refunds. While interesting, these negative amounts can muddy my analysis of donation amounts. I will separate these out into their own dataframe, as they do merit some analysis.
```{r}
AL <- subset(AL, AL$contb_receipt_amt > 0)
```

####Refund Analysis
Looking at refunds, I expect the most to be for Trump/Clinton, as these were the main candidates. 
```{r}
table(negAL$cand_nm)
```
To my surprise, Cruz had the most refunds, followed by Trump/Clinton. 


####Donation Histogram

```{r}
summary(AL$contb_receipt_amt)
```
```{r}
nrow(subset(AL,AL$contb_receipt_amt > 2000))
nrow(subset(AL,AL$contb_receipt_amt > 1000))
nrow(subset(AL,AL$contb_receipt_amt > 500))

```
### Candidate Donations
#### - Summary statistics
```{r}
#grouping by candidate name
candDonations <- group_by(AL, cand_nm)

#creating summary statistics
candDonations <- summarise(candDonations,count = n(),totalDonations = sum(contb_receipt_amt),meanDonation = mean(contb_receipt_amt),medianDonation = median(contb_receipt_amt))

#making it a data table then sorting by number of donations
candDonations <- data.table(candDonations)
candDonations <- candDonations[order(candDonations$count,decreasing = TRUE),]
candDonations
```

#### - Summary by Race
```{r}
#grouping by candidate name
electionDonations <- group_by(AL, election_tp)

#creating summary statistics
electionDonations <- summarise(electionDonations,count = n(),totalDonations = sum(contb_receipt_amt),meanDonation = mean(contb_receipt_amt),medianDonation = median(contb_receipt_amt))

#making it a data table then sorting by number of donations
electionDonations <- data.table(electionDonations)
electionDonations <- electionDonations[order(electionDonations$count,decreasing = TRUE),]
electionDonations
```
```{r}
#grouping by candidate name
candRaceDonations <- group_by(AL, cand_nm, election_tp)

#creating summary statistics
candRaceDonations <- summarise(candRaceDonations,count = n(),totalDonations = sum(contb_receipt_amt),meanDonation = mean(contb_receipt_amt),medianDonation = median(contb_receipt_amt))

#making it a data table then sorting by number of donations
candRaceDonations <- data.table(candRaceDonations)
candRaceDonations <- candRaceDonations[order(candRaceDonations$count,candRaceDonations$cand_nm,decreasing = TRUE),]
candRaceDonations
```

#### - Preparing time series analysis
```{r}
#grouping by date
donationsTime <- group_by(AL, contb_receipt_dt)

#creating summary statistics
donationsTime <- summarise(donationsTime,count = n(),totalDonations = sum(contb_receipt_amt),meanDonation = mean(contb_receipt_amt),medianDonation = median(contb_receipt_amt))

#making it a data table then sorting by number of donations
donationsTime <- data.table(donationsTime)
donationsTime <- donationsTime[order(donationsTime$contb_receipt_dt,decreasing = FALSE),]
donationsTime
```
#### How much money was raised per day?
```{r}
ggplot(donationsTime, aes(x = contb_receipt_dt, y = totalDonations)) + geom_line(color=I("blue")) + xlab("") + ylab("Daily Donation Sum") + scale_x_date(date_breaks = "4 month")
```
#### How many donations were there per day?
```{r}
ggplot(donationsTime, aes(x = contb_receipt_dt, y = count)) + geom_line(color=I("red")) + xlab("") + ylab("# Daily Donations")+scale_x_date(date_breaks = "4 month")
```
#### What was the mean and median donations per day?
##### - Mean
```{r}
ggplot(donationsTime, aes(x = contb_receipt_dt, y = meanDonation)) + geom_line(color=I("green")) + xlab("") + ylab("Mean of Daily Donations")+scale_x_date(date_breaks = "4 month")
```
##### - Median
```{r}
ggplot(donationsTime, aes(x = contb_receipt_dt, y = medianDonation)) + geom_line(color=I("orange")) + xlab("") + ylab("Median of Daily Donations")+scale_x_date(date_breaks = "4 month")
```

### Time Series Donation by Candidate
```{r}
#grouping by date
CandDonationsTime <- group_by(AL, contb_receipt_dt,cand_nm)

#creating summary statistics
CandDonationsTime <- summarise(CandDonationsTime,count = n(),totalDonations = sum(contb_receipt_amt),meanDonation = mean(contb_receipt_amt),medianDonation = median(contb_receipt_amt))

#making it a data table then sorting by number of donations
CandDonationsTime <- data.table(CandDonationsTime)
CandDonationsTime <- CandDonationsTime[order(CandDonationsTime$contb_receipt_dt,decreasing = FALSE),]
CandDonationsTime
```
#### How much money was raised per day?
```{r}
ggplot(CandDonationsTime, aes(x = contb_receipt_dt, y = totalDonations)) + geom_line() + xlab("") + ylab("Daily Donation Sum") + scale_x_date(date_breaks = "4 month")+facet_wrap(~cand_nm)
```

#### - What about the distribution for candidates with 1000+ donations?
For interetability purposes.

Filtering donations by candidate
```{r}
over1000 <- c("Clinton","Sanders","Cruz","Trump","Carson","Rubio")

filCndDon <- subset(CandDonationsTime, cand_nm %in% over1000)
```

```{r}
ggplot(filCndDon, aes(x = contb_receipt_dt, y = totalDonations,color=cand_nm)) + geom_line() + xlab("") + ylab("Daily Donation Sum") + scale_x_date(date_breaks = "9 month")+ylim(0,10000)+facet_wrap(~cand_nm)
```
#### # of Donations per major candidate?
```{r}
ggplot(filCndDon, aes(x = contb_receipt_dt, y = count ,color=cand_nm)) + geom_line() + xlab("") + ylab("# Daily Donations") + scale_x_date(date_breaks = "9 month")+facet_wrap(~cand_nm)+ylim(0,300)
```

#### - Median Donations per major candidate?
```{r}
ggplot(filCndDon, aes(x = contb_receipt_dt, y = medianDonation ,color=cand_nm)) + geom_line() + xlab("") + ylab("Median Daily Donations") + scale_x_date(date_breaks = "9 month")+facet_wrap(~cand_nm)
```

#### Donations over time for Democratic Candidates
```{r}
democrat <- c("Clinton","Sanders")
CandDonDemo <- subset(CandDonationsTime, cand_nm %in% democrat)

ggplot(CandDonDemo, aes(x = contb_receipt_dt, y = totalDonations,color=cand_nm)) + geom_line() + xlab("") + ylab("Sum of Daily Donations") + scale_x_date(date_breaks = "4 month")#+facet_wrap(~cand_nm)
```
```{r}
ggplot(CandDonDemo, aes(x = contb_receipt_dt, y = medianDonation,color=cand_nm)) + geom_line() + xlab("") + ylab("Median Daily Donations") + scale_x_date(date_breaks = "9 month")#+facet_wrap(~cand_nm)
```
#### - How many donations did each Democratic candidate recieve per day over time? 
```{r}
ggplot(CandDonDemo, aes(x = contb_receipt_dt, y = count,color=cand_nm)) + geom_line(aes(alpha=.7)) + xlab("") + ylab("# Daily Donations") + scale_x_date(date_breaks = "4 month")+ggtitle("Number of Donations to Democratic Candidates Over Time")
```


Perhaps a bit too grandular. Let's group by week.
```{r}
AL<- AL[order(AL$contb_receipt_dt,decreasing = FALSE)]
contb_wk <- data.table(Dates = AL$contb_receipt_dt, Week = format(AL$contb_receipt_dt, format = "%W"))

AL$contb_yr <- lapply(AL$contb_receipt_dt, year)

time("1999-01-01")
```


#### - What about distribution of all donations from major candidates?
Histogram Plot
```{r}
ggplot(filCndDon,aes(x = totalDonations,color=cand_nm,fill=I("white")))+
  geom_histogram(binwidth = 5) +coord_cartesian(xlim = c(0:100))#+facet_wrap(~cand_nm)
```



#### - What is the distribution of all donations?
```{r}
ggplot(AL,aes(x = contb_receipt_amt))+
  geom_histogram(binwidth = 5,fill=I("seagreen")) +coord_cartesian(xlim = c(0:100))

```
Donations are clustered at certain intervals (ex: 25,50,75,80,100), as one would expect. The largest number of values see, to be clustered around $25.

#### - What is the distribution of donations for each candidate?
```{r}
ggplot(AL,aes(x = contb_receipt_amt))+
  geom_histogram(binwidth = 5,fill=I("red")) +coord_cartesian(xlim = c(0:100))+facet_wrap(~cand_nm)

```

##Spatial Info
```{r}
#Committe ID, Candidate ID, and Candidate Name
head(sort(table(AL$contbr_city),decreasing = T),10)
head(sort(table(AL$contbr_zip),decreasing = T),10)
head(sort(table(AL$contbr_employer),decreasing = T),10)
head(sort(table(AL$contbr_occupation),decreasing = T),10)

```
What I aimed to do here was take a snapshot of the data and see how much it needed to be cleaned. The city, employer, and occupation vaules seem pretty well coded, but the zip codes (the 2nd line) will need be cleaned, via extraction of the first 5 values. 

For employer, joining similar values (Self, Self-Employed, Self Employed, etc.) will have to be done to get a more accurate snap-shot. I am not confident that there will be any useful insight from this variable, however.

For occupation, I will have to do minor cleaning (joining similar values), but this seems more stratified and varied. 

Because it is such a large group (30% of the dataset), it will be useful to analyze the spending patterns for retirees.


####Cleaning the Zip codes
```{r}
#Let's use the candidate's last name for interpretability in EDA
AL$contbr_zip <- substr(AL$contbr_zip,0,5)

```

